Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2015-04-01 paint-contour-tree.tex
abstract changes, example

\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
}
\date{}

@@

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first input-sensitive algorithm whose time complexity depends
on the contour tree struture, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to a linear time algorithm for contour tree computation for many
classes of natural instances. We complement with matching lower bounds for contour tree computation.

Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. 
There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

>>>>

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
}
\date{}

@@

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first input-sensitive algorithm whose time complexity depends
on the contour tree struture, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to a linear time algorithm for contour tree computation for many
classes of natural instances. 
%There is a crucial use of a family of binomial heaps to maintain priorities, 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.

Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. 
%There is a crucial use of a family of binomial heaps to 
%maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$. 
For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.
